Chatbots and messaging platforms in the classroom: an analysis from the teacher's perspective

### 2021-07-28 edubots_surveys.Rnw
Spin off part of the methodology to analysis
Mainly spell check

Several higher education chatbots' evaluation studies have been 
undertaken. For instance, a recent evaluation review study presented by Smutny and Schreiberova 
\cite{Smutny2020} examined 47 educational chatbots implemented in
Facebook Messenger with the focus to identify characteristics and
quality metrics such as language, subject matter and platform. 
Whereas study \cite{perez2020rediscovering} aimed to categorise educational chatbots according to their
purpose into service-oriented and teaching-oriented. The first category includes those that
provide service support such as the chatbot Ask Holly \cite{meetholly}
and Dina \cite{santoso2018dinus}; both chatbots respond to students'
questions about enrollment and registration. Ask L. U.
 \cite{LancasterUniversity_2019} answers students' frequently
asked questions about timetables, grades, tutors and societies. LISA
\cite{dibitonto2018chatbot} and Differ \cite{differ.chat} facilitate
breaking the ice between new students by introducing them to each
other. Ranoliya et al. \cite{ranoliya2017chatbot} proposed a generic chatbot for
university students that is able to answer to their frequently asked
questions. 
For some time, the University of Granada had ELVIRA as a
chatbot in its main web page \cite{MOREO20129}. Besides being able to
answer pre-established (frequent) questions, it did so from the website of the
University of Granada using an inset character, voiced by a real
person, who lip-synched the answers. The emphasis in the case of Elvira,
however, was on the authenticity of the synchronization more than in
making the update of questions and answers simpler, or more
interactive, using actual messaging platforms. 
In most cases, using the inset search engine was able to give you more up-to-date and
accurate answers than the ones provided by ELVIRA, which had to be updated by hand (and not too often).

@@

Despite its (arguably) successful implementation in different higher education institutions, 
over all in the fringes of educational activity, not in the actual interaction student-teacher, its implementation or deployment is not trivial.
Some authors, for instance \cite{sjostrom2018designing} have proposed a conceptual
architecture for the adoption of teaching-oriented chatbots in higher education.  This
conceptual architecture is based on a systematic literature review of
previous studies, examining the design of chatbots in higher education, as well as making a content 
analysis of student emails and discussion forum posts of
four instances of a Java programming course.  The study outlined
several design considerations; among them, the authors emphasised the
importance of developing chatbots in platforms that students and
educators are familiar with and can easily access (i.e. Facebook
Messenger) which was confirmed with \cite{hobert2019you} and
\cite{fernoagua2018intelligent} studies. In addition, Sj{\"o}str{\"o}m et al. \cite{sjostrom2018designing} 
argued that a conceptualision of learners' questions could aid designers in integrating the appropriate
types of questions that the chatbots should support for different
courses.  Coronado et al. \cite{coronado2018cognitive} proposed agents that store
learning materials to be provided on demand to students, whereas the authors in
\cite{crockett2017predicting} reported on tutoring systems,
which can perform initial assessments of students' understanding and
provide learning material that would advance their understanding to
the next level.

@@

Along this line of research, our previous work \cite{MoraChatbots2021} aimed at analysing the expectations of students on this regard; recent papers have also analyzed how the use of chatbots affected the learning outcomes of students in a Chinese class \cite{doi:10.1177/0735633120929622}. This paper analyzed conversational chatbots in an one-to-one setting, finding that learning really benefited from it. This is, once again, a proof of the benefit of chatbots in certain settings; however, there are some opinions that might prevent the successful acceptance of the technology, and these are what we are trying to find out in this paper, together with what kind of features would make its acceptance better.

Besides, the present work is focused on the other key actor in this challenge, i.e. educators. Their attitudes have been studied very recently by Chocarro et al. \cite{doi:10.1080/03055698.2020.1850426}, who, by analyzing surveys, created a {\em technology acceptance model} that proposed a series of features that would make chatbots easier to accept; among them, formality of language as well as easiness and usefulness. The survey targeted primary and secondary education teachers, and was also more interested in the general use of chatbots in education, not specifically in a classroom setting, seeking educational outcomes. However, they establish a series of results that are obviously interesting and relevant for this work.

@@

A total of 282 teachers responded to the survey: 193 teaching at the University (68.4\%) and 89 at other tertiary education institutions (31.6\%).

\textit{2. Which messaging app platform do you use to communicate with your peers or students?}
\begin{itemize}
\item Whatsapp
\item Telegram
\item LinkedIn
\item Twitter messages
\item Discord
\item Slack
\item Snapchat
\item The platform provided by your own institution
\item None, I do not use any messaging app to communicate with my students
\end{itemize}

\textit{2. Regarding the interaction with your students, you ...}
\begin{itemize}
\item Prefer it to be done by institutional means/tools: office hours, forum/messages via the virtual campus, institutional email. [\texttt{Officiality - Yes}]
\item Admit the use of other means/toosl: blogs, chat groups in messaging platforms, any others. [\texttt{Officiality - No}]
\end{itemize}

Responses were stored automatically in a Google Drive spreadsheet. Eventually, the results from the two forms we used for the first survey were collated in a single spreadsheet. Next we will show the results and analyze them.

@@

We will try to find the answers to the four research questions next, by analyzing the response to the different survey questions.

@@

Some of the questions in the survey were focused on the impact of the COVID19 pandemic between 2020 and 2021, in the teachers' viewpoints towards the use of instant messaging apps int their class. Our main intention in this case was to assess whether a crisis will bring about some kind of change in the use of tools.

@@

The majority of the teachers (n=165) mentioned that no changes in their habits occured due to the emergency remote teaching, as the use of messaging apps was part of their teaching practices and remains the same, from which 105 teachers come from the university, and 60 teachers from the vocational sector. The relationships between changes in the use of messaging apps, due to the covid-19 pandemic, and the instructors’ discipline, $X^{2}_{(24, N = 282)} = 44.856$, $p =0.006$, as well as the gender, $X^{2}_{(8, N = 282)} = 16.249$, p= 0.039 were also significant. This finding indicates that how teachers responded to the use of messaging apps during the pandemic was also related to their gender and discipline. In fact, from the majority of the teachers who did not change their habits in this respect (n=165), most of them are males (n=114), and come from the technology (n=60) and engineering (n=53) disciplines (see Tables \ref{tab:gender*covid}, \ref{tab:discipline*covid}).

@@

\begin{table}[h!tbp]
\centering
\resizebox{\textwidth}{!}{\begin{tabular}{lllllllll}
\hline
 & Connectivity & Familiarity  & Horizontallity & \begin{tabular}[c]{@{}l@{}}Official\\  formation\end{tabular} & Pluggability & Sustainability & Unidirectionality & Officiality \\ \hline
Analytics &  $12.172^{***}$ & $4.855^{*}$ & ns & $4.144^{*}$ &  $12.490^{***}$ & $4.314^{*}$ & ns & ns \\
Connectivity & - & ns & $5.971^{***}$ & ns & $16.286^{***}$ & ns & ns & ns \\
Hidden phone &  &  & ns & $5.678^{*}$ & $4.683^{*}$ & ns & $4.515^{*}$ & ns \\
Horizontallity &  &  - & ns & ns & ns & ns & ns & $5.453^{*}$ \\
\begin{tabular}[c]{@{}l@{}}Official\\  formation\end{tabular} &  & &  & - & $8.163^{**}$ & ns & ns & ns  \\
\end{tabular}}
\vspace{0.8em}
\caption{Interaction media features analysis ($^{***}p<0.001$, $^{**}p<0.01$, $^{*}p<0.05$). Columns and rows with no significant interaction have been suppressed for clarity. Every cell value represents $X^{2}_{(1, N = 205)}$. Details about each feature can be read in Section \ref{sec:meth}.}
\label{tab:interactionfeatures}
\end{table}

@@

Additionally, the question on COVID reflected the big challenge that represents changing habits, even in the case of a major crisis. Very few people actually changed, with women vs. men being an exception, as well as those teaching social sciences standing out among the rest. We do not have any kind of explanation for this behavior, other than technologies are perceived, and adopted, in a different way depending on background and, apparently, gender.

@@

Regarding gender grouping, both genders agree in their uses, so it seems not relevant at all. When looking at the plot containing the sector differentiation it is interesting ot observe how university teachers prefer in a higher proportion to use them for grades and FAQs in comparison with Vocational High School teachers. This seems reasonable as the frequency of seeing the students is smaller at the University so, automating the grade notification for continuous evaluation could seem much more comfortable using an automated tool than using the traditional platform. Furthermore, the high student/teacher ratios that university docents have to deal with could be another reason to have a higher number of them willing to use chatbots for FAQs. This could ameliorate the number of misunderstandings and email overload that so many students might generate.

@@

An analysis of survey questions related to RQ4 seems to indicate that teachers prefer to have interaction with their students in a chat "room"; a preference of {\em horizontal} interaction might indicate the need to delegate or offload part of the burden of answering, and attending, every single question posed by students. In an horizontal setting, other students can answer if the teacher does not do it immediately. This might also be the reason why the possibility of using tools from which analytics can be extracted is also valued by teachers: that way, activity in the chat tool, answering questions, regularity, can all be observed and valued as part of the process of assesing achievement of learning objectives by students.

@@

% What kidn of chatbots are preferred? (suggested by Fran)
Once that initial introduction is done, teachers (and students as well) will probably prefer the kind of bots that alleviate bureaucratic or repetitive tasks, such as answering frequent questions or answers on class or assignment schedules. These will help the introduction of more complex chatbots that will affect more directly learning outcomes, such as chatbots that help students integrate in the class, or are able to identify (and address) learning problems in students or in groups of them. These as well should be accompanied by analytics on student interaction, as well as possibly some natural language processing (in vernacular language) that will help assess the general mood of the class, and how different material (or external factors) affect it.

@@

% Only some COVID-response relationship.
In our survey, we couldn't find an answer to the right moment for doing this introduction of chatbot technologies. Response to questions related to change of behavior during and after the pandemic do not suggest that major (or minor) crisis could be an opportunity to introduce new technologies, since it does not bring major changes in attitude. A minor crisis would be, for instance, rollout of a new higher education law\footnote{We seem to have one of these, at least in Spain, every 10 years or so}, or introduction of new degrees. Although external changes do offer the chance of piloting new technologies, they do not seem to bring changes in attitudes in the teaching staff (which, after all, is bound to be the same). In absence of a clear answer in this direction, the right moment to introduce new technologies is always when the IT and managing staff is ready to support it (since, as we have seen before, "official" support is one of the factors that is most valued by teachers).

At any rate, the general feeling that transpires from the survey is that it is essential for any institution to take into account the stakeholder's opinion when introducing chatbots. This is true almost across the board, but in the case of chatbots (and instant messaging applications) their inmediacy and the fact that they can enter into what we could call the private sphere makes this even more necessary.

@@

% Emotinal, mood for future work... Chatbots perform sentiment analysis
Finally, the full extent of chatbot technology is not really being examined in these surveys. They can be connected to natural language processing engines with sentiment analysis as well as other analytics. Examining the mood of the class, as response to new material, assignment, exams or external events might help will help any student-centered teaching strategy, which will help students (and teachers) reach their learning objectives. This line of work, however, will be based on the introduction of some pilot study, combined with initial opinion assessment on the students.

@@

The datasets generated and analysed during the current study are available in the {\tt edubots-paper} repository, accesible from {\tt https://github.com/JJ/edubots-paper/tree/main/data}.

@@

>>>>

Several higher education chatbots' evaluation studies have been 
undertaken. For instance, a recent evaluation review study presented by Smutny and Schreiberova 
\cite{Smutny2020} examined 47 educational chatbots implemented in
Facebook Messenger with the focus to identify characteristics and
quality metrics such as language, subject matter and platform. 
Whereas study \cite{perez2020rediscovering} aimed to categorise educational chatbots according to their
purpose into service-oriented and teaching-oriented. The first category includes those that
provide service support such as the chatbot Ask Holly \cite{meetholly}
and Dina \cite{santoso2018dinus}; both chatbots respond to students'
questions about enrolment and registration. Ask L. U.
 \cite{LancasterUniversity_2019} answers students' frequently
asked questions about timetables, grades, tutors and societies. LISA
\cite{dibitonto2018chatbot} and Differ \cite{differ.chat} facilitate
breaking the ice between new students by introducing them to each
other. Ranoliya et al. \cite{ranoliya2017chatbot} proposed a generic chatbot for
university students that is able to answer to their frequently asked
questions. 
For some time, the University of Granada had ELVIRA as a
chatbot in its main web page \cite{MOREO20129}. Besides being able to
answer pre-established (frequent) questions, it did so from the website of the
University of Granada using an inset character, voiced by a real
person, who lip-synced the answers. The emphasis in the case of Elvira,
however, was on the authenticity of the synchronization more than in
making the update of questions and answers simpler, or more
interactive, using actual messaging platforms. 
In most cases, using the inset search engine was able to give you more up-to-date and
accurate answers than the ones provided by ELVIRA, which had to be updated by hand (and not too often).

@@

Despite its (arguably) successful implementation in different higher education institutions, 
over all in the fringes of educational activity, not in the actual interaction student-teacher, its implementation or deployment is not trivial.
Some authors, for instance \cite{sjostrom2018designing} have proposed a conceptual
architecture for the adoption of teaching-oriented chatbots in higher education.  This
conceptual architecture is based on a systematic literature review of
previous studies, examining the design of chatbots in higher education, as well as making a content 
analysis of student emails and discussion forum posts of
four instances of a Java programming course.  The study outlined
several design considerations; among them, the authors emphasised the
importance of developing chatbots in platforms that students and
educators are familiar with and can easily access (i.e. Facebook
Messenger) which was confirmed with \cite{hobert2019you} and
\cite{fernoagua2018intelligent} studies. In addition, Sj{\"o}str{\"o}m et al. \cite{sjostrom2018designing} 
argued that a conceptualisation of learners' questions could aid designers in integrating the appropriate
types of questions that the chatbots should support for different
courses.  Coronado et al. \cite{coronado2018cognitive} proposed agents that store
learning materials to be provided on demand to students, whereas the authors in
\cite{crockett2017predicting} reported on tutoring systems,
which can perform initial assessments of students' understanding and
provide learning material that would advance their understanding to
the next level.

@@

Along this line of research, our previous work \cite{MoraChatbots2021} aimed at analysing the expectations of students on this regard; recent papers have also analysed how the use of chatbots affected the learning outcomes of students in a Chinese class \cite{doi:10.1177/0735633120929622}. This paper analysed conversational chatbots in an one-to-one setting, finding that learning really benefited from it. This is, once again, a proof of the benefit of chatbots in certain settings; however, there are some opinions that might prevent the successful acceptance of the technology, and these are what we are trying to find out in this paper, together with what kind of features would make its acceptance better.

Besides, the present work is focused on the other key actor in this challenge, i.e. educators. Their attitudes have been studied very recently by Chocarro et al. \cite{doi:10.1080/03055698.2020.1850426}, who, by analysing surveys, created a {\em technology acceptance model} that proposed a series of features that would make chatbots easier to accept; among them, formality of language as well as easiness and usefulness. The survey targeted primary and secondary education teachers, and was also more interested in the general use of chatbots in education, not specifically in a classroom setting, seeking educational outcomes. However, they establish a series of results that are obviously interesting and relevant for this work.

@@

A total of 282 teachers responded to the survey: 193 teaching at the University (68.4\%) and 89 at other tertiary education institutions (31.6\%). Survey questions are shown in \ref{app:questions}.

Responses were stored automatically in a Google Drive spreadsheet. Eventually, the results from the two forms we used for the first survey were collated in a single spreadsheet. Next we will show the results and analyse them.

@@

We will try to find the answers to the four research questions next, by analysing the response to the different survey questions.

@@

Some of the questions in the survey were focused on the impact of the COVID19 pandemic between 2020 and 2021, in the teachers' viewpoints towards the use of instant messaging apps in their class. Our main intention in this case was to assess whether a crisis will bring about some kind of change in the use of tools.

@@

The majority of the teachers (n=165) mentioned that no changes in their habits occurred due to the emergency remote teaching, as the use of messaging apps was part of their teaching practices and remains the same, from which 105 teachers come from the university, and 60 teachers from the vocational sector. The relationships between changes in the use of messaging apps, due to the covid-19 pandemic, and the instructors’ discipline, $X^{2}_{(24, N = 282)} = 44.856$, $p =0.006$, as well as the gender, $X^{2}_{(8, N = 282)} = 16.249$, p= 0.039 were also significant. This finding indicates that how teachers responded to the use of messaging apps during the pandemic was also related to their gender and discipline. In fact, from the majority of the teachers who did not change their habits in this respect (n=165), most of them are males (n=114), and come from the technology (n=60) and engineering (n=53) disciplines (see Tables \ref{tab:gender*covid}, \ref{tab:discipline*covid}).

@@

\begin{table}[h!tbp]
\centering
\resizebox{\textwidth}{!}{\begin{tabular}{lllllllll}
\hline
 & Connectivity & Familiarity  & Horizontality & \begin{tabular}[c]{@{}l@{}}Official\\  formation\end{tabular} & Pluggability & Sustainability & Unidirectionality & Officiality \\ \hline
Analytics &  $12.172^{***}$ & $4.855^{*}$ & ns & $4.144^{*}$ &  $12.490^{***}$ & $4.314^{*}$ & ns & ns \\
Connectivity & - & ns & $5.971^{***}$ & ns & $16.286^{***}$ & ns & ns & ns \\
Hidden phone &  &  & ns & $5.678^{*}$ & $4.683^{*}$ & ns & $4.515^{*}$ & ns \\
Horizontality &  &  - & ns & ns & ns & ns & ns & $5.453^{*}$ \\
\begin{tabular}[c]{@{}l@{}}Official\\  formation\end{tabular} &  & &  & - & $8.163^{**}$ & ns & ns & ns  \\
\end{tabular}}
\vspace{0.8em}
\caption{Interaction media features analysis ($^{***}p<0.001$, $^{**}p<0.01$, $^{*}p<0.05$). Columns and rows with no significant interaction have been suppressed for clarity. Every cell value represents $X^{2}_{(1, N = 205)}$. Details about each feature can be read in Section \ref{sec:meth}.}
\label{tab:interactionfeatures}
\end{table}

@@

Additionally, the question on COVID reflected the big challenge that represents changing habits, even in the case of a major crisis. Very few people actually changed, with women vs. men being an exception, as well as those teaching social sciences standing out among the rest. We do not have any kind of explanation for this behaviour, other than technologies are perceived, and adopted, in a different way depending on background and, apparently, gender.

@@

Regarding gender grouping, both genders agree in their uses, so it seems not relevant at all. When looking at the plot containing the sector differentiation it is interesting to observe how university teachers prefer in a higher proportion to use them for grades and FAQs in comparison with Vocational High School teachers. This seems reasonable as the frequency of seeing the students is smaller at the University so, automating the grade notification for continuous evaluation could seem much more comfortable using an automated tool than using the traditional platform. Furthermore, the high student/teacher ratios that university teachers have to deal with could be another reason to have a higher number of them willing to use chatbots for FAQs. This could ameliorate the number of misunderstandings and email overload that so many students might generate.

@@

An analysis of survey questions related to RQ4 seems to indicate that teachers prefer to have interaction with their students in a chat "room"; a preference of {\em horizontal} interaction might indicate the need to delegate or offload part of the burden of answering, and attending, every single question posed by students. In an horizontal setting, other students can answer if the teacher does not do it immediately. This might also be the reason why the possibility of using tools from which analytics can be extracted is also valued by teachers: that way, activity in the chat tool, answering questions, regularity, can all be observed and valued as part of the process of assessing achievement of learning objectives by students.

@@

% What kind of chatbots are preferred? (suggested by Fran)
Once that initial introduction is done, teachers (and students as well) will probably prefer the kind of bots that alleviate bureaucratic or repetitive tasks, such as answering frequent questions or answers on class or assignment schedules. These will help the introduction of more complex chatbots that will affect more directly learning outcomes, such as chatbots that help students integrate in the class, or are able to identify (and address) learning problems in students or in groups of them. These as well should be accompanied by analytics on student interaction, as well as possibly some natural language processing (in vernacular language) that will help assess the general mood of the class, and how different material (or external factors) affect it.

@@

% Only some COVID-response relationship.
In our survey, we couldn't find an answer to the right moment for doing this introduction of chatbot technologies. Response to questions related to change of behaviour during and after the pandemic do not suggest that major (or minor) crisis could be an opportunity to introduce new technologies, since it does not bring major changes in attitude. A minor crisis would be, for instance, rollout of a new higher education law\footnote{We seem to have one of these, at least in Spain, every 10 years or so}, or introduction of new degrees. Although external changes do offer the chance of piloting new technologies, they do not seem to bring changes in attitudes in the teaching staff (which, after all, is bound to be the same). In absence of a clear answer in this direction, the right moment to introduce new technologies is always when the IT and managing staff is ready to support it (since, as we have seen before, "official" support is one of the factors that is most valued by teachers).

At any rate, the general feeling that transpires from the survey is that it is essential for any institution to take into account the shareholder's opinion when introducing chatbots. This is true almost across the board, but in the case of chatbots (and instant messaging applications) their immediacy and the fact that they can enter into what we could call the private sphere makes this even more necessary.

@@

Finally, the full extent of chatbot technology is not really being examined in these surveys. They can be connected to natural language processing engines with sentiment analysis as well as other analytics. Examining the mood of the class, as response to new material, assignment, exams or external events might help will help any student-centred teaching strategy, which will help students (and teachers) reach their learning objectives. This line of work, however, will be based on the introduction of some pilot study, combined with initial opinion assessment on the students.

@@

The datasets generated and analysed during the current study are available in the {\tt edubots-paper} repository, accessible from {\tt https://github.com/JJ/edubots-paper/tree/main/data}.

@@

\section{Appendix I: Survey questions}
\label{app:questions}

\textit{2. Which messaging app platform do you use to communicate with your peers or students?}
\begin{itemize}
\item WhatsApp
\item Telegram
\item LinkedIn
\item Twitter messages
\item Discord
\item Slack
\item Snapchat
\item The platform provided by your own institution
\item None, I do not use any messaging app to communicate with my students
\end{itemize}

\textit{2. Regarding the interaction with your students, you ...}
\begin{itemize}
\item Prefer it to be done by institutional means/tools: office hours, forum/messages via the virtual campus, institutional email. [\texttt{Officiality - Yes}]
\item Admit the use of other means/tools: blogs, chat groups in messaging platforms, any others. [\texttt{Officiality - No}]
\end{itemize}
